### 2016-03-02 CICM2016/appendices.tex
More changes

We use a simplified syntax for the GHC Core language, given below in BNF style ($[]$ and $(,)$ denote repetition and grouping, respectively). We mechanically translate the full GHC Core language into this sub-set, which does not contain types, prior to performing recurrent clustering.

%\begin{figure}
  \begin{equation}\label{fig:coresyntax}
    \begin{split}
      expr\    \rightarrow\ & \CVar\ id                          \\
                         |\ & \CLit\ literal                     \\
                         |\ & \CApp\ expr\ expr                  \\
                         |\ & \CLam\ \mathcal{L}\ expr           \\
                         |\ & \CLet\ bind\ expr                  \\
                         |\ & \CCase\ expr\ \mathcal{L}\ [alt]   \\
                         |\ & \CType                             \\
      id\      \rightarrow\ & \CLocal\       \mathcal{L}         \\
                         |\ & \CGlobal\      \mathcal{G}         \\
                         |\ & \CConstructor\ \mathcal{D}         \\
      literal\ \rightarrow\ & \CLitNum\ \mathcal{N}              \\
                         |\ & \CLitStr\ \mathcal{S}              \\
      alt\     \rightarrow\ & \CAlt\ altcon\ expr\ [\mathcal{L}] \\
      altcon\  \rightarrow\ & \CDataAlt\ \mathcal{D}             \\
                         |\ & \CLitAlt\ literal                  \\
                         |\ & \CDefault                          \\
      bind\    \rightarrow\ & \CNonRec\ binder                   \\
                         |\ & \CRec\ [binder]                    \\
      binder   \rightarrow\ & \CBind\ \mathcal{L}\ expr
    \end{split}
  \end{equation}
  Where:
  \begin{tabular}[t]{l @{ $=$ } l}
    $\mathcal{S}$ & string literals    \\
    $\mathcal{N}$ & numeric literals   \\
    $\mathcal{L}$ & local identifiers  \\
    $\mathcal{G}$ & global identifiers \\
    $\mathcal{D}$ & constructor identifiers
  \end{tabular}
%\end{figure}

@@

The recursive definition is mostly routine; each repeated element (shown as $\dots$) has an example to indicate their handling, e.g. for $\CRec$ we apply $toTree$ to each $e_i$. We ignore values of $\mathcal{D}$, since constructors have no internal structure for us to compare; they can only be compared based on their types, which we do not currently support. We also ignore values from $\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$, and we conjecture that the effect on clustering real code is low.

>>>>

The GHC Core language is based on \fc{}, and described in detail in \cite[Appendix C]{sulzmann2007system}. For our machine learning purposes we are mostly interested in the syntax of reducible expressions (representing Haskell of the form \hs{f a b \dots = \dots}), and use the simplified syntax given below in BNF style ($[]$ and $(,)$ denote repetition and grouping, respectively):

\begin{equation}\label{fig:coresyntax}
  \begin{split}
    expr\    \rightarrow\ & \CVar\ id                          \\
                       |\ & \CLit\ literal                     \\
                       |\ & \CApp\ expr\ expr                  \\
                       |\ & \CLam\ \mathcal{L}\ expr           \\
                       |\ & \CLet\ bind\ expr                  \\
                       |\ & \CCase\ expr\ \mathcal{L}\ [alt]   \\
                       |\ & \CType                             \\
    id\      \rightarrow\ & \CLocal\       \mathcal{L}         \\
                       |\ & \CGlobal\      \mathcal{G}         \\
                       |\ & \CConstructor\ \mathcal{D}         \\
    literal\ \rightarrow\ & \CLitNum\ \mathcal{N}              \\
                       |\ & \CLitStr\ \mathcal{S}              \\
    alt\     \rightarrow\ & \CAlt\ altcon\ expr\ [\mathcal{L}] \\
    altcon\  \rightarrow\ & \CDataAlt\ \mathcal{D}             \\
                       |\ & \CLitAlt\ literal                  \\
                       |\ & \CDefault                          \\
    bind\    \rightarrow\ & \CNonRec\ binder                   \\
                       |\ & \CRec\ [binder]                    \\
    binder   \rightarrow\ & \CBind\ \mathcal{L}\ expr
  \end{split}
\end{equation}
Where:
\begin{tabular}[t]{l @{ $=$ } l}
  $\mathcal{S}$ & string literals    \\
  $\mathcal{N}$ & numeric literals   \\
  $\mathcal{L}$ & local identifiers  \\
  $\mathcal{G}$ & global identifiers \\
  $\mathcal{D}$ & constructor identifiers
\end{tabular}

The full Core language emitted by GHC (as of version 7.10.2, the latest at the time of writing) is translated automatically to this simplified form prior to recurrent clustering. Our major restriction is to ignore type-level entities (such as datatype definitions, explicit casts, and differences between types/kinds/coercions). Our implementation also handles several other forms of literal (machine words of various sizes, individual characters, etc.), but we omit them here for brevity as their treatment is similar to those of strings and numerals.

@@

The recursive definition is mostly routine; each repeated element (shown as $\dots$) has an example to indicate their handling, e.g. for $\CRec$ we apply $toTree$ to each $e_i$. We ignore values of $\mathcal{D}$, since constructors have no internal structure for us to compare; they can only be compared based on their types, which we do not currently support. We also ignore values from $\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$, and we conjecture that the effect of such ``magic values'' on clustering real code is low.

### 2016-03-02 CICM2016/background.tex
More changes

Like most functional programming languages, Haskell builds upon $\lambda$-calculus, with extra features such as a strong type system and ``syntactic sugar'' to improve readability. For simplicity, we will focus on an intermediate representation of the \textsc{GHC} compiler, known as \emph{GHC Core}, rather than the relatively large and complex syntax of Haskell proper. Core is based on \fc{}, but for our machine learning purposes we are mostly interested in its syntax; for a more thorough treatment of \fc{} and its use in GHC, see \cite[Appendix C]{sulzmann2007system}.

The sub-set of Core we consider is shown in Figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major restriction is to ignore type hints (such as explicit casts, and differences between types/kinds/coercions). For brevity, we also omit several other forms of literal (machine words of various sizes, individual characters, etc.), as their treatment is similar to those of strings and numerals. We will use quoted strings to denote names and literals, e.g. \texttt{Local "foo"}, \texttt{Global "bar"}, \texttt{Constructor "Baz"}, \texttt{LitStr "quux"} and \texttt{LitNum "42"}, and require only that they can be compared for equality.

\begin{figure}
  \begin{haskell}\begin{verbatim}
data Nat = Z
         | S Nat

@@

even :: Nat -> Bool
even    Z  = True
even (S n) = odd n\end{verbatim}
  \end{haskell}
  \caption{A Haskell datatype for Peano numerals with some simple arithmetic functions, including mutually-recursive definitions for \hs{odd} and \hs{even}. \hs{Bool} is Haskell's built in boolean type, which can be regarded as \hs{data Bool \equal{} True | False}.}
  \label{fig:haskellexample}
\end{figure}

\begin{figure}
  \begin{small}
    \underline{\texttt{odd}}
    \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              (Alt (DataAlt "Z") (Var (Constructor "False")))
              (Alt (DataAlt "S") (App (Var (Global "even"))
                                      (Var (Local  "n")))
                                 "n"))
    \end{verbatim}

    \underline{\texttt{even}}
    \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              (Alt (DataAlt "Z") (Var (Constructor "True")))
              (Alt (DataAlt "S") (App (Var (Global "odd"))
                                      (Var (Local  "n")))
                                 "n"))
    \end{verbatim}
  \end{small}
  \caption{Translations of functions in Figure \ref{fig:haskellexample} into the Core syntax of Figure \ref{fig:coresyntax}. Notice the introduction of explicit $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case} to represent piecewise definitions. Fresh variables are chosen arbitrarily as \texttt{"a"}, \texttt{"b"}, etc.}
  \label{fig:coreexample}
\end{figure}

Figure \ref{fig:haskellexample} shows some simple Haskell function definitions, along with a custom datatype for Peano numerals. The translation to our Core syntax is routine, and shown in Figure \ref{fig:coreexample}. Although the Core is more verbose, we can see that similar structure in the Haskell definitions gives rise to similar structure in the Core; for example, the definitions of \hs{odd} and \hs{even} are identical in both languages, except for the particular identifiers used. It is this close correspondence which allows us to analyse Core expressions in place of their more complicated Haskell source.

Note that we exclude representations for type-level entities, including datatype definitions like that of \hs{Nat}. GHC can represent these, but in this work we only consider reducible expressions (i.e. value-level bindings of the form \mbox{\hs{f a b ... = ...}}).

@@

Such conjectures can be used in several ways: they can be simplified for direct presentation to the user (by removing any equation which can be derived from the others by rewriting), sent to a more rigorous system like \hspec{} for proving, or even serve as a background theory for an automated theorem prover \cite{claessen2013automating}.

@@

Although complete, this enumeration approach is wasteful: many terms are unlikely to appear in theorems, which requires careful choice by the user of what to include in the signature. For example, we know that addition and multiplication are closely related, and hence obey many algebraic laws. Our machine learning technique aims to predict these kinds of relations between functions, so we can create small signatures which nevertheless have the potential to give rise to many equations.

@@

Our approach to scaling up \qspec{} takes inspiration from two sources. The first is relevance filtering, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs. We describe this approach in more details in \S \ref{sec:relevance}. Relevance filtering is a practical tool which has existing applications in software, such as the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.

Despite the idea's promise, we cannot simply invoke existing relevance filter algorithms in our theory exploration setting. The reason is that relevance filtering is a supervised learning method, i.e. it would require a distinguished expression to compare everything against. Theory exploration does not have such a distinguished expression; instead, we are interested in relationships between \emph{all} terms in a signature, and hence we must consider the relevance of \emph{all terms} to \emph{all other terms}.

@@

We use \emph{k-means} clustering, implemented in the Weka tool \cite{Holmes.Donkin.Witten:1994} by Lloyd's algorithm \cite{lloyd1982least}, as this setup is used by ML4PG and ACL2(ml). For simplicity, we will cluster $n$ points into $k = \lceil \sqrt{\frac{n}{2}} \rceil$ clusters, following the ``rule of thumb'' given in \cite[pp. 365]{mardia1979multivariate}, and will use randomly-selected elements of our data set to determine the initial clusters (known as the Forgy method).

>>>>

Like most functional programming languages, Haskell builds upon $\lambda$-calculus, with extra features such as a strong type system and ``syntactic sugar'' to improve readability. To avoid the relatively large and complex syntax of Haskell proper, we will focus on an intermediate representation of the \textsc{GHC} compiler known as \emph{GHC Core}, described in Appendix \ref{sec:core}.

We will use quoted strings to denote names and literals, e.g. \texttt{Local "foo"}, \texttt{Global "bar"}, \texttt{Constructor "Baz"}, \texttt{LitStr "quux"} and \texttt{LitNum "42"}, and require only that they can be compared for equality.

\begin{figure}
  \begin{haskell}\begin{verbatim}
data Bool = True | False

data Nat = Z | S Nat

not :: Bool -> Bool
not True  = False
not False = True

@@

even :: Nat -> Bool
even    Z  = True
even (S n) = odd n\end{verbatim}
  \end{haskell}
  \caption{Haskell datatypes for booleans and natural numbers, followed by some simple function definitions. Note that \hs{odd} and \hs{even} are mutually recursive.}
  \label{fig:haskellexample}
\end{figure}

\begin{figure}
  \begin{small}
    \underline{\texttt{not}}
    \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              (Alt (DataAlt "True")  (Var (Constructor "False")))
              (Alt (DataAlt "False") (Var (Constructor "True"))))
    \end{verbatim}

    \underline{\texttt{odd}}
    \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              (Alt (DataAlt "Z") (Var (Constructor "False")))
              (Alt (DataAlt "S") (App (Var (Global "even"))
                                      (Var (Local  "n")))
                                 "n"))
    \end{verbatim}

    \underline{\texttt{even}}
    \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              (Alt (DataAlt "Z") (Var (Constructor "True")))
              (Alt (DataAlt "S") (App (Var (Global "odd"))
                                      (Var (Local  "n")))
                                 "n"))
    \end{verbatim}
  \end{small}
  \caption{Translations of functions in Figure \ref{fig:haskellexample} into our Core syntax, with fresh variables chosen arbitrarily as \texttt{"a"}, \texttt{"b"}, etc. Notice the introduction of explicit $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case} to represent piecewise definitions. The \texttt{"b"} variables are introduced to preserve sharing.}
  \label{fig:coreexample}
\end{figure}

Figure \ref{fig:haskellexample} shows some simple Haskell function definitions, along with common definitions for boolean and natural number datatypes (analogous to those in Haskell's standard library). The translation to our Core syntax is routine, and shown in Figure \ref{fig:coreexample}. Although Core is more verbose, we can see that similar structure in the Haskell definitions gives rise to similar structure in the Core; for example, the definitions of \hs{odd} and \hs{even} are identical in both languages, except for the particular identifiers used. It is this close correspondence which allows us to analyse Core expressions in place of their more complicated Haskell source.

@@

Such conjectures can be used in several ways: they can be presented to the user, sent to a more rigorous system like \hspec{} for proving, or even serve as a background theory for an automated theorem prover \cite{claessen2013automating}. As an example, giving the definitions from Figure \ref{fig:haskellexample} to \qspec{}. along with some suitable \hs{Nat} and \hs{Bool} variables, produces equations including \hs{not (not a) == a} and \hs{not (odd x) == even x}.

@@

Although complete, this enumeration approach is wasteful: many terms are unlikely to appear in theorems, which requires careful choice by the user of what to include in the signature. For example, we know that addition and multiplication are closely related, and hence obey many algebraic laws. Our machine learning technique aims to predict these kinds of relations between functions, so we can create small signatures which can be explored quickly, yet have the potential to give rise to many equations.

@@

Our approach to scaling up \qspec{} takes inspiration from two sources. The first is relevance filtering, which makes expensive algorithms used in theorem proving more practical by only considering clauses deemed ``relevant'' to the problem \cite{meng2009lightweight}.

Relevance is determined by comparing clauses to the target theorem, but theory exploration does not have such a distinguished term. Instead, we are interested in relationships between \emph{all} terms in a signature, and hence we need a different algorithm for considering the relevance of \emph{all terms} to \emph{all other terms}.

@@

Due to its use by ML4PG and ACL2(ml), we use \emph{k-means} clustering, implemented in the Weka tool \cite{Holmes.Donkin.Witten:1994} by Lloyd's algorithm \cite{lloyd1982least}, with randomly-selected input elements as the initial clusters. Rather than relying on the user to provide the number of clusters $k$, we use the ``rule of thumb'' given in \cite[pp. 365]{mardia1979multivariate} of clustering $n$ data points into $k = \lceil \sqrt{\frac{n}{2}} \rceil$ clusters.

### 2016-03-02 CICM2016/future.tex
More changes

The most obvious next step for our algorithm is to incorporate types. Types contain valuable information about an expression, and allow us to distinguish between constructors. Our algorithm closely follows that of ML4PG, which \emph{does} support types, by populating matrix cells with tokens \emph{and} their types. Unfortunately this is more complicated in Haskell than it is in Coq, since types form a separate part of the language from terms, and we do not have an interactive Core environment to query for types (unlike ML4PG, which runs inside the Proof General environment).

We can also compare the performance of our hand-selected features with \emph{learned} representations, like those reviewed in \cite{bengio2013representation}. This may provide an indication of how important it is to understand the language when identifying salient aspects of expressions, and how difficult various aspects of it might be to learn.

>>>>

The most obvious next step is to incorporate types. Types contain valuable information about an expression, and would allow us to distinguish between constructors. Since our algorithm closely follows that of ML4PG, which \emph{does} support types, the only barrier is the practical issue of propagating annotations through all of the Core definitions.

More speculative directions include the use of \emph{learned} representations, rather than our hand-crafted features \cite{bengio2013representation}. This would provide an interesting comparison, as well as being more robust in the face of language evolution. Another intriguing possibility is to extend the recurrent nature of our algorithm to make use of the discovered properties during clustering; for example, by treating the discovered equations as rewrite rules to reduce the ASTs prior to feature extraction.

### 2016-03-02 CICM2016/implementation.tex
More changes

A post-processing stage determines which Core definitions can be used by \qspec{}, based on their type, visibility (whether they are encapsulated inside their module or visible to importers), etc. For each definition, we simply use the GHCi interpreter to check if calling \qspec{} with that argument is well-typed.

These definitions are then sorted topologically, based on the non-local identifiers appearing in their ASTs, and feature vectors are constructed using a Haskell implementation of the approach described in \S \ref{sec:recurrentclustering}. Since the features associated with each identifier may vary between iterations, we leave the raw identifiers in the vector so their features can be extracted in the correct context.

The iterative algorithm \ref{alg:recurrent} is a simple shell script, which calls the Weka machine learning system to perform k-means clustering and associate each Core definition with a cluster number. These numbers are used to finish the deferred feature extraction of identifiers, the resulting feature vectors are clustered, and so on until all SCCs have been processed.

Once the recurrent clustering is complete, we split the definitions based on their final clusters, and use these clusters to generate strings of Haskell code for invoking \qspec{}. To construct a \qspec{} signature from these definitions, we must:

\begin{itemize}
  \item{Monomorphise}: If a value has polymorphic type, e.g. \hs{equal :: forall t. t -> t -> Bool}, we must choose a concrete representation to use (in this case for \hs{t}), in order to know which random generator to use. We take the approach used in \qcheck{} by attempting to instantiate all type variables to \hs{Integer}. Any values where this is invalid, e.g. if there are incompatible class constraints (e.g. \hs{forall t. IsString t => t}, where \hs{Integer} does not implement \hs{IsString}) will have already been discarded in our post-processing stage after extracting the ASTs.

  \item{Qualification}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity. There is still the possibility that multiple packages will declare modules of the same name, although this is rare as it causes problems for any Haskell programmer trying to use those modules. In such cases the exploration process simply aborts.

  \item{Variable definition}: Once a \qspec{} theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \qspec{}'s exhaustive search).

@@

>>>>

A post-processing stage determines which Core definitions can be used by \qspec{}, based on their type, visibility (whether they are encapsulated inside their module or visible to importers), etc. For each definition, we simply use the GHCi interpreter to check if calling \qspec{} with that argument is well-typed. This information, along with type signatures, arity, etc. are stored alongside the Core definitions in JSON format.

The definitions are then sorted topologically, based on the non-local identifiers appearing in their ASTs, and feature vectors are constructed using a Haskell implementation of the approach described in \S \ref{sec:recurrentclustering}. Since the features associated with each identifier may vary between iterations (as the clusters change), we leave the raw identifiers in the vector so their features can be extracted in the correct context.

We implement Algorithm \ref{alg:recurrent} as a simple shell script, which invokes Weka for clustering and associates each Core definition with a cluster number. These numbers are used to finish the deferred feature extraction of identifiers, the resulting feature vectors are clustered, and the process is repeated until all SCCs have been processed.

Once the recurrent clustering is complete, we generate a string of Haskell code for each of the final clusters, which will invoke \qspec{} with a suitable signature. This involves:

\begin{itemize}
  \item{Monomorphising}: If a value has polymorphic type, e.g. \hs{equal :: forall t. t -> t -> Bool}, we must choose a concrete representation to use (in this case for \hs{t}), in order to know which random generator to use. We take the approach used in \qcheck{} by attempting to instantiate all type variables to \hs{Integer}. Any values where this is invalid, such as those with incompatible class constraints (e.g. \hs{forall t. IsString t => t}, where \hs{Integer} does not implement \hs{IsString}) will not be included in the signature (this is checked at the AST post-processing stage).

  \item{Qualifying}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity. There is still the possibility that multiple packages will declare modules of the same name, although this is rare as it causes problems for any Haskell programmer trying to use those modules. In such cases the exploration process simply aborts.

  \item{Appending variables}: Once a \qspec{} theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \qspec{}'s exhaustive search).

@@

The equations resulting from evaluating these strings are collected and outputted as JSON values, to ease further processing (e.g. displaying in some form, sending to a theorem prover, etc.).

### 2016-03-02 CICM2016/intro.tex
More changes

Functional programming claims to be more ``mathematically tractable'' than other ``conventional'' approaches, due to properties like \emph{referential transparency} and \emph{immutability}, which allow equational reasoning without many of the complications caused by side effects and strict evaluation order \cite{hughes1989functional}. Languages such as Haskell combine these advantages with static type systems to restrict the possible values of every expression, allowing powerful analysis tools such as property checkers \cite{claessen2011quickcheck} and theorem provers \cite{rosen2012proving} to be developed.

>>>>

Functional programming claims to be more ``mathematically tractable'' than other ``conventional'' approaches, due to properties like \emph{referential transparency} and \emph{immutability}, which allow equational reasoning without many of the complications caused by side effects and strict evaluation order \cite{hughes1989functional}. Languages such as Haskell combine these advantages with static type systems to restrict the possible values of every expression, making programs amenable to powerful analysis tools such as property checking \cite{claessen2011quickcheck} and theorem proving \cite{rosen2012proving}.

### 2016-03-02 CICM2016/related.tex
More changes

We briefly described theory exploration in \S \ref{sec:theoryexploration}, as the task of discovering \emph{new} theorems in a software or proof library, rather than proving/disproving user-provided statements. This general idea is quite old, with . The current generation of theory exploration tools trace their roots back to the interactive \textsc{Theorema} \cite{buchberger2000theory} system of Buchberger. In such a setting, most of the automation focuses on bookkeeping (such as type-checking, managing scope, etc.) and more traditional automated theorem proving.

Subsequent systems have increased the level of automation, and have shown promise in tasks such as lemma discovery. As well as \qspec{} and \hspec{} in Haskell, automated theory exploration has been applied to Isabelle \cite{Montano-Rivas.McCasland.Dixon.ea:2012} \cite{johansson2009isacosy} \cite{Hipster}. Since \qspec{} is used for conjecture generation in many of those systems, they can also benefit from our work with machine learning.

@@

The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \cite{haken1985intractability}. The idea of \emph{relevance filtering} (or, \emph{premise selection}) is to reduce the input size of a task by removing redundant or irrelevant information from the input. Our restriction of theory exploration to intelligently-selected clusters of symbols, rather than whole libraries at a time, is the essentially the same idea being applied in a novel setting.

Relevance filtering has mostly been used in automated proof search, where it simplifies problems by removing from consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed \emph{irrelevant}. The technique is used in Isabelle's Sledgehammer tool, during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included. This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a clause is relevant: how do we know what will be required, before we have the proof?

Various approaches have been proposed in the literature \cite{kuhlwein2013mash} \cite{kuhlwein2012overview} \cite{alama2014premise}, though the original approach is based on the proportion of symbols which a clause has in common with the problem statement \cite{meng2009lightweight} (plus various heuristics).

@@

Our recurrent clustering approach takes inspiration from the ML4PG \cite{journals/corr/abs-1212-3618} and ACL2(ml) \cite{heras2013proof} tools, used for analysing proofs in Coq and ACL2, respectively. Whilst both transform syntax trees into matrices, the algorithm of ML4PG most closely resembles ours as it assigns tokens directly to matrix elements. In contrast, the matrices produced by ACL2(ml) \emph{summarise} information about the tree; for example, one column counts the number of variables appearing at each tree level, others count the number of function symbols which are nullary, unary, binary, etc. Whilst it may be interesting to contrast our current algorithm with an alternative based on that of ACL2(ml), it is unclear how such summaries could be extended to include types, which seems the next logical step for our approach. The ML4PG algorithm extends trivially, by using (term, type) pairs instead of just terms.

@@

ACL2(ml) reasons by analogy: finding theorem statements which are similar to the current goal, and attempting to prove the goal in a similar way. In particular, the lemmas used to prove a theorem are mutated by substituting symbols for those which appear in the same cluster. For example, if \texttt{plus} and \texttt{multiply} are clustered together, and we are trying to prove a goal involving \texttt{multiply}, then ACL2(ml) might consider an existing theorem involving \texttt{plus}. The lemmas used to prove that theorem will be mutated, for example replacing occurrences of \texttt{plus} with \texttt{mult}, in an attempt to prove the goal.

Whilst we do not currently reason by analogy, this is an interesting area for future work in theory exploration: given a set of theorems relating particular terms, we might form conjectures regarding similar terms found through clustering.

>>>>

We briefly described theory exploration in \S \ref{sec:theoryexploration}, as the task of discovering \emph{new} theorems in a software or proof library, rather than proving/disproving user-provided statements. This general idea dates back at least to the 1970s, most famously to work by Lenat \cite{lenat1977automated}, though the theory exploration tools used in functional programming trace their roots back to the interactive \textsc{Theorema} \cite{buchberger2000theory} system of Buchberger.

As well as \qspec{} and \hspec{} in Haskell, automated theory exploration has been applied to Isabelle \cite{Montano-Rivas.McCasland.Dixon.ea:2012} \cite{johansson2009isacosy} \cite{Hipster}. Since \qspec{} is used for conjecture generation in many of these systems, they may also benefit from our machine learning approach.

@@

The idea of \emph{relevance filtering} (or, \emph{premise selection}) is to reduce the search space of a theorem proving task by removing redundant or irrelevant information (axioms, definitions, lemmas, etc.) from the input. Our restriction of theory exploration to intelligently-selected clusters of symbols, rather than whole libraries at a time, is essentially the same idea applied in a novel setting.

The technique is used in Isabelle's Sledgehammer tool, during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included. This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a clause is relevant: how do we know what will be required, before we have the proof?

The original approach uses the proportion of symbols which a clause has in common with the problem statement \cite{meng2009lightweight} (plus various heuristics), although various alternatives have been proposed in the literature \cite{kuhlwein2013mash} \cite{kuhlwein2012overview} \cite{alama2014premise}.

@@

Our recurrent clustering algorithm is similar to that of ML4PG, as our transformation maps the elements of a syntax tree to distinct cells in a matrix. In contrast, the matrices produced by ACL2(ml) \emph{summarise} the tree elements: providing, for each level of the tree, the number of variables, nullary symbols, unary symbols, etc.

Unlike ML4PG, our algorithm can handle mutually-recursive definitions, but it also ignores types. This is because obtaining the type of each component of a Core definition is more difficult than in ML4PG (which queries the current \textsc{Proof General} state), and is left for future work.

@@
